---
title: "Cancer Analysis"
author: "Caleb Aguiar"
date: "2/27/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)#For summary stats
library(ggplot2)#For graphs
library(scales)#Clean up axis labels
library(corrplot)#Explore correlations
library(gplots)#For plotting means
library(caTools)
library(rpart)
library(rpart.plot)
library(factoextra)
library(cluster)
library(caTools)
library(tidyverse)

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

# Introduction

This introduction was paraphrased from the Breast Cancer Surveillance Consortium overview of the dataset. The purpose of analysis of this dataset is to determine casual factors related with breast cancer and practice applying statistical methodology. The following analysis concerns a dataset of 280,660 mammograms from women in the Breast Cancer Surveillance Consortium. All women did not have a previous diagnosis of breast cancer and did not have any imaging in the previous nine months preceding the mammogram. Cancer registry and pathology data wwere linked to the mammography data.


The dataset is intended to be used for risk estimation. We will use different methods to predict risk based on the variables that we may have. We will see how accurate our model is with just demographics, then a model with basic patient history(which a patient themselves would know and be able to answer in a questionnaire, the application of this maybe a web application where you input your information and get your risk), and then finally all variables including ones that a medical professional may only be able to interpret or generate(this could be in the form of a application that a medical professional would be able to input measured variables based on a visit and calculate a paticular patient's risk). 


Data collection and sharing was supported by the National Cancer Institute-funded Breast Cancer Surveillance Consortium (HHSN261201100031C). You can learn more about the BCSC at: http://www.bcsc-research.org/.

```{r, eval = TRUE} 
#Import
dt <- read.delim("risk.txt", header = FALSE, sep = " ", dec = ".")
#Remove extra cols
dt<- dt[c(1,3:16)]
#Rename cols
colnames(dt)<-c('menopaus','agegrp','density','race','Hispanic','bmi','agefirst','nrelbc','brstproc','lastmamm','surgmeno','hrt','invasive','cancer','training')
head(dt)
tail(dt)
```
```{r}
describe(dt)
```


```{r echo=FALSE}
dt$agegrp<-as.factor(dt$agegrp)
dt$bmi<-as.factor(dt$bmi)
dt$race<-as.factor(dt$race)
dt$Hispanic<-as.factor(dt$Hispanic)
```

# Demographics

```{r echo=FALSE, warning=FALSE}
ages<-c('35-39','40-44','45-49','50-59','60-64','65-69','70-74','75-79','80-84')
bmival<-c('NA','10-24.99','25-29.99','30-34.99','35+','unknown')
race<-c('NA','white','asian','black','nat ame','mix','unk')
his<-c('no','yes','unk','unk','.','unk')

p1 <- ggplot(dt, aes(x=agegrp)) +
    geom_bar(fill="powderblue") +
    ggtitle("Age Groups") + theme(axis.text.x = element_text(angle = 65, hjust = 1))+ scale_x_discrete(labels=c(ages))

p2 <- ggplot(dt, aes(x=race)) +
    geom_bar(fill="powderblue") +
    ggtitle("Race")+ theme(axis.text.x = element_text(angle = 65, hjust = 1))+ scale_x_discrete(labels=c(race))

p3 <- ggplot(dt, aes(x=Hispanic)) +
    geom_bar(fill="powderblue") +
    ggtitle("Hispanic")+ theme(axis.text.x = element_text(angle = 65, hjust = 1))+ scale_x_discrete(labels=c(his))

p4 <- ggplot(dt, aes(x=bmi)) +
    geom_bar(fill="powderblue") +
    ggtitle("BMI")+ theme(axis.text.x = element_text(angle = 65, hjust = 1))+ scale_x_discrete(labels=c(bmival))       

multiplot(p1,p2,p3,p4,cols=2)
```
When we look at some of the demographics for the data, we see that we have a right skew for age groups, with demographic data that is/is not representative of the population.  

```{r echo=FALSE}
dt$agegrp<-as.integer(dt$agegrp)
dt$bmi<-as.integer(dt$bmi)
dt$race<-as.integer(dt$race)
dt$Hispanic<-as.integer(dt$Hispanic)
corrplot(cor(dt[1:14]))
```
There doesn't seem to be many strong correlations other than "invasive" and "cancer".

# K-Means

```{r echo=FALSE, warning=FALSE}
set.seed(1234)
dt.no.na<-na.omit(dt)
dt.sc <- scale(dt.no.na)
describe(dt.sc)

dt.pc<-prcomp(dt.sc)
summary(dt.pc)

```

```{r}
dt.pc$rotation
```

```{r}
plot(dt.pc, type='l')
```



```{r}
KmeansClustering <- kmeans(dt.sc, centers = 7)

KmeansClustering$size
fviz_cluster(KmeansClustering, data = dt.sc)

```
```{r}
biplot(dt.pc)
abline(h = 0, v = 0, col = "gray60")
```

```{r}

dt$cancer<-as.factor(dt$cancer)
colors<-c("green","red")
plot(dt.pc$x,pch=20,col=colors[dt$cancer])

```

```{r}
library(xgboost)
#Import
dt <- read.delim("risk.txt", header = FALSE, sep = " ", dec = ".")
#Remove extra cols
dt<- dt[c(1,3:16)]
#Rename cols
colnames(dt)<-c('menopaus','agegrp','density','race','Hispanic','bmi','agefirst','nrelbc','brstproc','lastmamm','surgmeno','hrt','invasive','cancer','training')
unique(dt$cancer)
```

```{r}
split = sample.split(dt$cancer, SplitRatio = 0.7)
Train = subset(dt, split==TRUE)
Test = subset(dt, split==FALSE)
#Look at different abilities to predict based on what values we have

#Only demographics
demo.col<-c(2,4,5,6)#Only agegroup, race, and bmi

hist.col<-c(1,2,3,4,5,6,7,8,9)#without testing, add patient history to demographics

med.col<-c(1:13)


```


```{r}

xgb.cancer.demo<- xgboost(data = data.matrix(Train[demo.col]), 
                  label = Train[,14], 
                  eta = 0.1,
                  max_depth = 1, 
                  nround=100, 
                  subsample = 1,
                  colsample_bytree = 1,
                  num_class = 1,
                  min_child_weight = 10,
                  gamma = 7,
                  nthread = 30,
                  eval_metric = "logloss",
                  objective = "binary:logistic",
                  verbose = 0
                  )

xgb.cancer.hist<- xgboost(data = data.matrix(Train[hist.col]), 
                  label = Train[,14], 
                  eta = 0.1,
                  max_depth = 1, 
                  nround=100, 
                  subsample = 1,
                  colsample_bytree = 1,
                  num_class = 1,
                  min_child_weight = 10,
                  gamma = 7,
                  nthread = 30,
                  eval_metric = "logloss",
                  objective = "binary:logistic",
                  verbose = 0
                  )

xgb.cancer.med<- xgboost(data = data.matrix(Train[med.col]), 
                  label = Train[,14], 
                  eta = 0.1,
                  max_depth = 1, 
                  nround=100, 
                  subsample = 1,
                  colsample_bytree = 1,
                  num_class = 1,
                  min_child_weight = 10,
                  gamma = 7,
                  nthread = 30,
                  eval_metric = "logloss",
                  objective = "binary:logistic",
                  verbose = 0
                  )
```





```{r}
Test$xgb.pred.demo <- predict(xgb.cancer.demo, data.matrix(Test[demo.col]))
Test$xgb.pred.hist<- predict(xgb.cancer.hist, data.matrix(Test[hist.col]))
Test$xgb.pred.med<- predict(xgb.cancer.med, data.matrix(Test[med.col]))
```



```{r}
library(pROC)
auc.demo = roc(Test$cancer,Test$xgb.pred.demo)
auc_curve.xgb.demo <- plot.roc(Test$cancer, Test$xgb.pred.demo, main = "ROC curve Demographics XGBoost", percent=TRUE, cex.main=0.75, cex.lab = 0.75)
auc(auc.demo)

```
```{r}
auc.hist = roc(Test$cancer,Test$xgb.pred.hist)
auc_curve.xgb.hist <- plot.roc(Test$cancer, Test$xgb.pred.hist, main = "ROC curve Patient History XGBoost", percent=TRUE, cex.main=0.75, cex.lab = 0.75)
auc(auc.hist)

```

```{r}
auc.med = roc(Test$cancer,Test$xgb.pred.med)
auc_curve.xgb.med <- plot.roc(Test$cancer, Test$xgb.pred.med, main = "ROC curve Medical Exam History XGBoost", percent=TRUE, cex.main=0.75, cex.lab = 0.75)
auc(auc.med)

```



# Introduction

Breast Cancer Wisconsin(Diagnostic) Data Set

Predict whether the cancer is benign or malignant

Features are computed from a digitized image of a fine needle aspirate (FNA) of a breast mass. They describe characteristics of the cell nuclei present in the image.

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

```{r, eval = TRUE} 
#Import
dt.wi <- read.csv("data.csv")
dt.wi<-dt.wi[,1:32]
dt.wi$diagnosis<-as.factor(dt.wi$diagnosis)

head(dt.wi)
tail(dt.wi)
```
```{r}
describe(dt.wi)
```
```{r}
dt.wi.sc <- scale(dt.wi[3:10])
describe(dt.wi.sc)

dt.wi.pc<-prcomp(dt.wi.sc)
summary(dt.wi.pc)
```
```{r}
dt.wi.pc$rotation

```

```{r}
plot(dt.wi.pc, type='l')

```
```{r}
biplot(dt.wi.pc,xlim=c(-.25,.15), ylim=c(-.12,.12))
abline(h = 0, v = 0, col = "gray60")
```

```{r}
plot(dt.wi.pc$x,pch=20,col=dt.wi$diagnosis)
```




```{r}
KmeansClustering <- kmeans(dt.wi.sc, centers = 2)

fviz_cluster(KmeansClustering, data = dt.wi.sc)
```




```{r}
unique(dt.wi$X)
```



```{r}
cancer.p<-table(dt.wi[2])
cancer.p
percent.malignant<-paste(round(cancer.p[[2]]/(cancer.p[[1]]+cancer.p[[2]])*100,digits = 3), "%")
percent.malignant
```
```{r}

split = sample.split(dt.wi$diagnosis, SplitRatio = 0.7)
Train = subset(dt.wi, split==TRUE)
Test = subset(dt.wi, split==FALSE)

# Building a Logistic Regression Model
dt.wi.log = glm(diagnosis~., data = dt.wi, family = binomial, na.action = na.pass)
summary(dt.wi.log)

head(dt.wi)

```







glm.fit: algorithm did not converge glm.fit: fitted probabilities numerically 0 or 1 occurred
Issue with model, going to dig into what is happening by using a tree

There is one fairly common circumstance in which both convergence problems and the Hauck-Donner phenomenon can occur. This is when the fitted probabilities are extremely close to zero or one. Consider a medical diagnosis problem with thousands of cases and around 50 binary explanatory variable (which may arise from coding fewer categorical variables); one of these indicators is rarely true but always indicates that the disease is present. Then the fitted probabilities of cases with that indicator should be one, which can only be achieved by taking βi = ∞. The result from glm will be warnings and an estimated coefficient of around +/- 10. There has been fairly extensive discussion of this in the statistical literature, usually claiming non-existence of maximum likelihood estimates; see Sautner and Duffy (1989, p. 234).

Read more about this error here:https://stackoverflow.com/questions/8596160/why-am-i-getting-algorithm-did-not-converge-and-fitted-prob-numerically-0-or







```{r}

CancerTree = rpart(outcome ~ ., method="class", data=dt.wi, minbucket=1)
prp(CancerTree)
```







```{r}
dt.log = glm(cancer~., data = Train, family = binomial)
summary(dt.log)


```
